---
title: "Exploratory Data Analysis (EDA) - AmeriCorps"
author: "DATS 6101 Group8 [Anusha Umashankar](https://github.com/Anusha-raju)| [Aidan Carlisle](https://github.com/acarlisle8)| [Rachel Thomas](https://github.com/releered)| [Sayan Patra](https://github.com/Sayanpatraa)"
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
# Once installed, load the library.
library(ezids)
```


```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
options(scientific=T, digits = 3) 
```

```{r include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
```



# Introduction on AmeriCorps

[AmeriCorps](https://americorps.gov/) is a national service program in the United States, established in 1993, that encourages individuals to serve their communities through various initiatives. Members engage in projects that address pressing social issues such as education, health, public safety, and environmental sustainability. In return for their service, members may receive benefits like education awards, living stipends, and professional experience.


AmeriCorps Seniors is a specific branch of the AmeriCorps program aimed at older adults, typically aged 55 and over. It provides opportunities for seniors to contribute their skills and experience to community service. Programs under AmeriCorps Seniors include:

  * Senior Corps: This encompasses several programs like Foster Grandparents, Senior Companions, and RSVP (Retired and Senior Volunteer Program), allowing seniors to engage in meaningful volunteer work that often supports younger generations or helps those in need.

  * Benefits: Participants in AmeriCorps Seniors can receive stipends, training, and the chance to make a positive impact in their communities while also combating social isolation.

Together, AmeriCorps and AmeriCorps Seniors play a crucial role in fostering civic engagement and addressing community needs across the country.



There are several programs in AmeriCorps:

* AmeriCorps

  * AmeriCorps NCCC
  
  * AmeriCorps State and National
  
  * AmeriCorps VISTA
  
  * Public Health AmeriCorps
  
* AmeriCorps Seniors

  * AmeriCorps Seniors Foster Grandparent Program
  
  * AmeriCorps Seniors RSVP
  
  * AmeriCorps Seniors Senior Companion Program
  

# Data Source

* The [Data](https://catalog.data.gov/dataset/americorps-participant-demographics-data) is collected and published by AmeriCorps.
* The dataset provides comparisons of demographic group prevalence in AmeriCorps Member/Volunteers populations to that of the greater U.S. population. 
* The data was initially published on October 19, 2023, and last modified on December 18, 2023.




# Info on the dataset

```{r}
americorps <- read.csv("/home/anusha/workspace/Data science/AmeriCorps-Volunteers/AmeriCorps_Participant_Demographics_Data.csv", header =TRUE)
```
* The source dataset consists of about 26000 observations over 17 attributes 
* The data includes the following columns:
```{r}
names(americorps)
```

* Of these, for the analysis we are only interested in the following
```{r}
americorps <- americorps %>%
  select(-odds_ratio_estimate,-lwr_95,-uppr_95, -representation,-magnitude_range_size, -significant)
names(americorps)
```

* The data has already been cleared of missing or NA values and is being converted to desired variable type.

The summary of the data:
```{r}
americorps$geography<-as.factor(americorps$geography)
americorps$dem_cat<-as.factor(americorps$dem_cat)
americorps$dem_group<-as.factor(americorps$dem_group)
americorps$program_office<-as.factor(americorps$program_office)
americorps$service_region<-as.factor(americorps$service_region)
americorps$state<-as.factor(americorps$state)
americorps$time_period<-as.factor(americorps$time_period)
summary(americorps)
```

# Preprocessing the data

* The summary rows of the analysis, including program office categories such as "All AmeriCorps Members," "AmeriCorps Member Programs (no NCCC)," and "AmeriCorps Seniors (All)," were also included in the dataset; therefore, they are removed.

```{r}
americorps_subset <- americorps %>%
  filter(!program_office %in% c("All AmeriCorps Members", "AmeriCorps Member Programs (no NCCC)", "AmeriCorps Seniors (All)"))
```

* The data also had summary rows in the Demographic Group category such as "Racially Diverse (Non-White)"; therefore, they are removed. 
```{r}
americorps_subset <- americorps_subset %>% filter(!dem_group %in% c("Racially Diverse (Non-White)"))
```

* For this analysis we are only considering the data with respect to "Race" demographic analysis specifically in 2022.
```{r}
dem_cat_subset <- 'Race'
americorps_subset <- americorps_subset %>%
  filter(dem_cat == dem_cat_subset)

time_period_subset <- '2022'
americorps_subset <- americorps_subset %>%
  filter(time_period == time_period_subset)
```

* The data is being recorded at three geographic levels (National, State, Region). To avoid data duplication, the data is being considered at one particular level at a time.

```{r}
#filter by national
americorps_national <- americorps_subset %>%
  filter(geography == "National")
#filter by state
americorps_state <- americorps_subset %>%
  filter(geography == "State")
#filter by region
americorps_region <- americorps_subset %>%
  filter(geography == "Region")

```


# Key points regarding the dataset

* The total count of volunteers in a particular demograph is not measured as the number of volunteers in the demograph, rather is the summed number of Member Service Years (MSYs) or AmeriCorps Seniors volunteers belonging to the specific demographic group of interest.

# SMART Questions

**1. What is the racial composition of the volunteers?**
```{r}
group_MSY_national <- aggregate(count_dem_group ~ dem_group, data = americorps_national, sum)
total_MSY_national <- sum(group_MSY_national$count_dem_group)
group_MSY_national$percentage <- (group_MSY_national$count_dem_group / total_MSY_national) * 100

ggplot(group_MSY_national, aes(x = dem_group, y = percentage, fill = dem_group)) +
  geom_bar(stat = "identity", width = 0.6) +
  labs(title = "Race Distribution by Percentage of Member Service Years (MSY)", x = "Race", y = "MSY(%)") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") + 
  geom_text(aes(label = round(percentage, 1)), vjust = 0, size = 3) +
  theme(axis.text.x = element_blank()) +
  theme(legend.position = "none",
        plot.title = element_text(size = 12, face = "bold", hjust = 0),
        axis.title = element_text(size = 10, face = "bold"),
        axis.text.x = element_text(size = 6, angle = 90, vjust = -3),
        axix.text.y = element_text(size = 10))+
  scale_y_continuous(breaks = seq(0, 100, by = 15))
```

## Notes:

1. This graph describes the racial distribution by percentage of Member Service Years (MSY).
2. The graph above displays racial groups along the x-axis and MSY percentages on the y-axis.
3. The graph clearly indicates that volunteers identifying as white represent the highest percentage within the volunteer program.
4. However, this observation does not suggest that other groups are not participating in volunteering activities. Instead, it indicates that the representation of these groups within the overall population reflects a similar odds ratio. This means that while certain groups may have higher visibility or recognition in volunteering efforts, others are equally active, but their engagement is proportionate to their demographic presence. Therefore, the data highlights a correlation between population distribution and volunteering rates, rather than an absence of participation from less represented groups.


**1. Which program offices are most effective in engaging a diverse range of participants?**
```{r}
```

**2. How does the representation of different demographic groups compare to their proportions in the general population?**
```{r}
```
**3. Which groups are overrepresented or underrepresented in AmeriCorps?**
```{r}
```
**4. How do participation rates compare between different programs within the same service region?**
```{r}
```

**6.Are some racial groups more engaged in volunteer activities within specific states or program offices?**
```{r}
```
