---
title: "Exploratory Data Analysis (EDA) - AmeriCorps"
author: "GWU Intro to Data Science -- DATS 6101 Group8"
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
# Once installed, load the library.
library(ezids)
```


```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
options(scientific=T, digits = 3) 
```

```{r include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
```



# Introduction on AmeriCorps

AmeriCorps is a national service program in the United States, established in 1993, that encourages individuals to serve their communities through various initiatives. Members engage in projects that address pressing social issues such as education, health, public safety, and environmental sustainability. In return for their service, members may receive benefits like education awards, living stipends, and professional experience.


AmeriCorps Seniors is a specific branch of the AmeriCorps program aimed at older adults, typically aged 55 and over. It provides opportunities for seniors to contribute their skills and experience to community service. Programs under AmeriCorps Seniors include:

  * Senior Corps: This encompasses several programs like Foster Grandparents, Senior Companions, and RSVP (Retired and Senior Volunteer Program), allowing seniors to engage in meaningful volunteer work that often supports younger generations or helps those in need.

  * Benefits: Participants in AmeriCorps Seniors can receive stipends, training, and the chance to make a positive impact in their communities while also combating social isolation.

Together, AmeriCorps and AmeriCorps Seniors play a crucial role in fostering civic engagement and addressing community needs across the country.


[AmeriCorps]: https://americorps.gov/


There are several programs in AmeriCorps:

* AmeriCorps

  * AmeriCorps NCCC
  
  * AmeriCorps State and National
  
  * AmeriCorps VISTA
  
  * Public Health AmeriCorps
  
* AmeriCorps Seniors

  * AmeriCorps Seniors Foster Grandparent Program
  
  * AmeriCorps Seniors RSVP
  
  * AmeriCorps Seniors Senior Companion Program
  

# Data Source

* The data is collected and published by AmeriCorps.
* The dataset provides comparisons of demographic group prevalence in AmeriCorps Member/Volunteers populations to that of the greater U.S. population. 
* The data was initially published on October 19, 2023, and last modified on December 18, 2023.

[Data]: https://catalog.data.gov/dataset/americorps-participant-demographics-data


# Info on the dataset

```{r}
americorps <- read.csv("/home/anusha/workspace/Data science/AmeriCorps-Volunteers/AmeriCorps_Participant_Demographics_Data.csv", header =TRUE)
```
* The source dataset consists of about 26000 observations over 17 attributes 
* The data includes the following columns:
```{r}
names(americorps)
```

* Of these, for the analysis we are only interested in the following
```{r}
americorps <- americorps %>%
  select(-odds_ratio_estimate,-lwr_95,-uppr_95, -representation,-magnitude_range_size, -significant)
names(americorps)
```

* The data has already been cleared of missing or NA values and is being convert to desired variable type.

The summary of the data:
```{r}
americorps$geography<-as.factor(americorps$geography)
americorps$dem_cat<-as.factor(americorps$dem_cat)
americorps$dem_group<-as.factor(americorps$dem_group)
americorps$program_office<-as.factor(americorps$program_office)
americorps$service_region<-as.factor(americorps$service_region)
americorps$state<-as.factor(americorps$state)
americorps$time_period<-as.factor(americorps$time_period)
summary(americorps)
```


```{r}
dem_cat_subset <- 'Race'
amco <- americorps %>%
  filter(dem_cat == dem_cat_subset)

time_period_subset <- '2022'
amco <- amco %>%
  filter(time_period == time_period_subset)

amco <- amco %>%
  filter(!program_office %in% c("All AmeriCorps Members", "AmeriCorps Member Programs (no NCCC)", "AmeriCorps Seniors (All)"))

amco <- amco %>%
  filter(!dem_group %in% c("Racially Diverse (Non-White)"))
```


**1. Which program offices are most effective in engaging a diverse range of participants?**
```{r}
national <- amco %>%
  filter(geography == "National")
print(national)

race_program_national <- aggregate(count_dem_group ~ program_office + dem_group, data = national, sum)
print(race_program_national)

program_race_national <- national %>%
  group_by(program_office, dem_group) %>%
  summarise (total_dem_group = sum(count_dem_group))
print(program_race_national)

ggplot(program_race_national, aes(x = program_office, y = total_dem_group, fill = dem_group)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Demographic Groups by Program Office", 
       x = "Program Office", 
       y = "Total MSY") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

**2. How does the representation of different demographic groups compare to their proportions in the general population?**
```{r}
```
**3. Which groups are overrepresented or underrepresented in AmeriCorps?**
```{r}
```
**4. How do participation rates compare between different programs within the same service region?**
```{r}
region <- amco_r2022_noNW %>%
  filter(geography == "Region")
print(region)

program_region <- region %>%
  group_by(service_region, program_office) %>%
  summarise (total_dem_group = sum(count_dem_group))
print(program_region)

ggplot(program_region, aes(x = service_region, y = total_dem_group, fill = program_office)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Demographic Groups by Program Office and Service Region", 
       x = "Service Region", 
       y = "Total MSY") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

Overall impression is RSVP has a significantly higher level of participation than the rest of the groups, and the rest seem somewhat similar.
```{r}
mean_progMSY_national <- national %>%
  group_by(program_office) %>%
  summarise(mean_dem_group = mean(count_dem_group, na.rm = TRUE))  

print(mean_progMSY_national)
```
H0: Mean MSY is equal between programs
H1: Mean MSY is not equal between programs
```{r}
anova_progMSY <- aov(count_dem_group ~ program_office, data = national)
summary(anova_progMSY)
```
p = 0.00822 < 0.05
Reject H0
```{r}
tukey_progMSY <- TukeyHSD(anova_progMSY)
print(tukey_progMSY)
```
p<0.05 for RSVP vs NCCC, RSVP vs SCP, RSVP vs SDP, RSVP vs VISTA, RSVP vs PH indicating RSVP mean MSY is statistically different from mean MSY in those programs. Remainder are >0.05 indiating their means are statistically equal.

**5. What is the racial composition of the volunteers?**
```{r}
national <- amco %>%
  filter(geography == "National")
print(national)

program_MSY_national <- national %>%
  group_by(program_office) %>%
  summarise (total_dem_group = sum(count_dem_group))
print(program_MSY_national)

race_program_national <- aggregate(count_dem_group ~ program_office + dem_group, data = national, sum)
print(race_program_national)

group_MSY_national <- aggregate(count_dem_group ~ dem_group, data = national, sum)
total_MSY_national <- sum(group_MSY_national$count_dem_group)
group_MSY_national$percentage <- (group_MSY_national$count_dem_group / total_MSY_national) * 100
print(group_MSY_national)


#bar chart w %MSY
ggplot(group_MSY_national, aes(x = dem_group, y = percentage, fill = dem_group)) +
  geom_bar(stat = "identity", width = 0.6) +
  labs(title = "Race Distribution by Percentage of Member Service Years (MSY)", x = "Race", y = "MSY(%)") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") + 
  geom_text(aes(label = round(percentage, 1)), vjust = 0, size = 3) +
  theme(axis.text.x = element_blank()) +
  theme(legend.position = "none",
        plot.title = element_text(size = 12, face = "bold", hjust = 0),
        axis.title = element_text(size = 10, face = "bold"),
        axis.text.x = element_text(size = 6, angle = 90, vjust = -3),
        axix.text.y = element_text(size = 10))+
  scale_y_continuous(breaks = seq(0, 100, by = 15))
```
**6.Are some racial groups more engaged in volunteer activities within specific states or program offices?**

By state
```{r}
state <- amco %>%
  filter(geography == "State")
print(state)

race_state <- state %>%
  group_by(state, dem_group) %>%
  summarise (total_dem_group = sum(count_dem_group))
print(race_state)

#stacked bar chart
ggplot(race_state, aes(x = state, y = total_dem_group, fill = dem_group)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Demographic Groups by State", x = "State", y = "Total MSY") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

By Program Office
```{r]}
program_race_national <- national %>%
  group_by(program_office, dem_group) %>%
  summarise (total_dem_group = sum(count_dem_group))
print(program_race_national)

ggplot(program_race_national, aes(x = program_office, y = total_dem_group, fill = dem_group)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Demographic Groups by Program Office", 
       x = "Program Office", 
       y = "Total MSY") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```